Master Thesis

### 2013-06-13 the-last-part.tex
Buildup to the definition of GCGs

The types we have been showing up to now propose a very straightforward
notion of subtyping. All of the types can be seen as conjunctions of
properties where $\_$ (underscore) is the conjunction operator. A type
$\tau$ can then be seen as a subtype of type $\sigma$ whenever the
conjuncts of $\tau$ form a superset of the conjuncts of $\sigma$. This
is not unlike the structural typing of records in OCaml (CITE SOME OCAML
SPEC HERE), where a record type is a subtype of another record type if
the former provides, among others, all the fields provided by the latter
and gives them the same type. In our case, one type is a subtype of
another if the former gives, among others, all the properties given by
the latter. This notion of subtyping can be expressed by adding the rule
of inference in Figure~\ref{fig:type-superset} to our type
system.\footnote{In the rule }

@@

@@

$$
S_{tatou} : N \land NEG{=}F \land NUM{=}SG
$$
$$
S_{le_{NEG_1}} : (N \land NEG{=}F) \limp (((NP \land NEG{=}F) \limp S) \limp S)
$$
$$
S_{le_{NEG_2}} : (N \land NEG{=}T) \limp (((NP \land NEG{=}T) \limp S) \limp S)
$$
$$
S_{le_{NUM}} : (N \land NUM{=}SG) \limp (((NP \land NEG{=}SG) \limp S) \limp S)
$$
$$
S_{le_{VAR}} : N \limp (((NP \land VAR{=}F) \limp S) \limp S)
$$

>>>>

The types we have been showing up to now propose a very straightforward
notion of subtyping. All of the types can be seen as conjunctions of
properties where $\_$ (underscore) is the conjunction operator. A type
$\tau$ can then be seen as a subtype of type $\sigma$ whenever the
conjuncts of $\tau$ form a superset of the conjuncts of $\sigma$. This
is not unlike the structural typing of records in OCaml (CITE SOME OCAML
SPEC HERE), where a record type is a subtype of another record type if
the former provides, among others, all the fields provided by the latter
and gives them the same type. In our case, one type is a subtype of
another if the former gives, among others, all the properties given by
the latter. This notion of subtyping can be expressed by adding the rule
of inference in Figure~\ref{fig:type-superset} to our type
system.\footnote{In the rule \ldots (talk about $\land$ instead of $\_$
  (underscore) here...)}

@@

(Talk about whether or not this is a conservative extension...)

@@

$$
S_{tatou} : N \land NEG{=}F \land NUM{=}SG
$$
$$
S_{le_{NEG_1}} : (N \land NEG{=}F) \limp (((NP \land NEG{=}F) \limp S) \limp S)
$$
$$
S_{le_{NEG_2}} : (N \land NEG{=}T) \limp (((NP \land NEG{=}T) \limp S) \limp S)
$$
$$
S_{le_{NUM}} : (N \land NUM{=}SG) \limp (((NP \land NUM{=}SG) \limp S) \limp S)
$$
$$
S_{le_{VAR}} : N \limp (((NP \land VAR{=}F) \limp S) \limp S)
$$

\ldots

\subsection{Intersections of Languages}

We will now turn to another solution to the complexity of grammars
implementing multiple independent mechanisms.

When ACGs are discussed, one of the oft-praised properties is
compositionality. Let us consider a situation where we have three
signatures $\Sigma_1$, $\Sigma_2$ and $\Sigma_3$, their respective
distinguished types $S_1$, $S_2$ and $S_3$ and two lexicons
$\mathcal{L}_{12} : \Sigma_1 \to \Sigma_2$ and $\mathcal{L}_{23} :
\Sigma_2 \to \Sigma_3$ such that $\mathcal{L}_{12}(S_1) = S_2$ and
$\mathcal{L}_{23}(S_2) = S_3$. (TODO: INCLUDE A DIAGRAM OF THIS
SITUATION)

We can now define languages $A_1$, $A_2$ and $A_3$ where $A_i$ is the
set of terms from $\Lambda(\Sigma_i)$ that have the type $S_i$. Such
languages, which are defined only in terms of a signature and some
distinguished type, are called \emph{abstract languages} in the
terminology of ACGs. They consist of terms whose structure is
constrained only by the types assigned by the signature.

Further than that, we can look at the languages $\mathcal{L}_{12}(A_1)$
and $\mathcal{L}_{23}(A_2)$, images of the abstract languages given by
the lexicons. Since the lexicons are homomorphic mappings w.r.t. the
types, all elements of $\mathcal{L}_{12}(A_1)$ will have the type
$\mathcal{L}_{12}(S_1) = S_2$ and will therefore also belong to $A_2$
(similarly for $\mathcal{L}_{23}(A_2)$ and $A_3$). These new languages,
which are formed by mapping an ACG abstract language using a lexicon,
are called \emph{object languages}. Their terms are constrained not only
by the types of the signature they are built upon but also by the type
system of the abstract signature.

Finally, we can also take the composition of lexicons $\mathcal{L}_{13}
= \mathcal{L}_{23} \circ \mathcal{L}_{12}$, which itself is also a
lexicon, and map it over $A_1$. This gives us a language
$\mathcal{L}_{13}(A_1) = \mathcal{L}_{23}(\mathcal{L}_{12}(A_1))$ which
is at the same time the image of the object language
$\mathcal{L}_{12}(A_1)$ given by $\mathcal{L}_{23}$ and also the image
of the abstract language $A_1$ given by $\mathcal{L}_{13}$. This points
to an interesting property of ACG languages: they are closed under
transformation by a lexicon.

In the ACG literature, ACGs are defined as quadruples $\mathcal{G} =
\mathopen{<}\Sigma_A, \Sigma_O, \mathcal{L}, S\mathclose{>}$ where
$\Sigma_A$ is the abstract signature, $\Sigma_O$ is the object
signature, $\mathcal{L} : \Sigma_A \to \Sigma_O$ is a lexicon linking
the two and $S$ is a distinguished type of the abstract signature
$\Sigma_A$. From this grammar, we can define two languages. The abstract
language $\mathcal{A}(\mathcal{G}) = \{t \in \Lambda(\Sigma_A) \mid
\ \vdash_{\Sigma_A} t : S\}$ and the object language
$\mathcal{O}(\mathcal{G}) = \mathcal{L}(\mathcal{A}(\mathcal{G})) = \{t
\in \Lambda(\Sigma_O) \mid \exists u \in \mathcal{A}(\mathcal{G}),
\mathcal{L}(u) = t\}$. The composition of two ACGs $\mathcal{G}_1 =
\mathopen{<}\Sigma_1, \Sigma_2, \mathcal{L}_{12}, S_1\mathclose{>}$ and
$\mathcal{G}_2 = \mathopen{<}\Sigma_2, \Sigma_3, \mathcal{L}_{23},
S_2\mathclose{>}$ is defined as $\mathcal{G}_2 \circ \mathcal{G}_1 =
\mathopen{<}\Sigma_1, \Sigma_3, \mathcal{L}_{23} \circ \mathcal{L}_{12},
S_1\mathclose{>}$.

The definition of ACGs given above is quite practical in that it is
complete, it gives a precise account of what set of languages are
generated by ACGs (abstract and object languages for which there exist
the quadruples described above). This contrasts with the presentation of
ACGs that we gave in~\ref{ssec:acg}, where we defined ACGs as
collections of signatures with associated distinguished types and
connected with lexicons. Our definition only introduced the requisite
machinery, but delayed the delimitation of the way languages can be
defined. Leaving this question open will let us answer it now by
providing an extension/generalization of ACGs which will help us
overcome the modularity issues that have been the theme of this
section.\footnote{Incidentally, this was not the reason for the way we
  presented ACGs in~\ref{ssec:acg}. We chose this style as it grouped
  together the information in a more natural way. We approach signatures
  as descriptions of domains (strings, syntax, semantics), each one
  being associated with a distinguished type telling us which objects
  the domain set out to describe (strings, sentences, truth
  values). Having distinguished types directly attached to signatures
  allows us to omit them from the ACG quadruples. However, this only
  leaves us with triples containing the two signatures and the lexicon
  connecting them, and since we can infer the signatures themselves from
  the domain and the range of the lexicon, we do not have much
  motivation to introduce these tuples. Instead of composing ACGs, we
  compose just the lexicons.

On top of that, we will want to define signatures and lexicons that
yield different object languages (at least some string representation of
the form and some semantic representation of the meaning), which would
mean having at least two different ACGs which are linked (e.g. by
sharing the same abstract signature). However, since these two grammars
are very closely related, we would prefer to reason about our grammar as
\emph{one} system/graph of signatures and lexicons which defines
multiple languages instead of talking about \emph{multiple} grammars
which share some similarities.}

Before we turn to our proposal, we will quickly recall some of the other
styles of composition used in ACGs (a more detailed exposition can be
found in \cite{pogodalla2012controlling}). Besides making the object
signature of one grammar be the abstract signature of another as we saw
above (TODO: INSERT REFERENCE TO FUTURE DIAGRAM), we can have two
grammars share the same abstract signature allowing us to transduce
between terms in the two object signatures. See Figure~(TODO: INSERT A
DIAGRAM FOR THIS SITUATION) for a diagrammatic example.

These two modes of composition can be fruitfully mixed. Consider (TODO:
INSERT A DIAGRAM), where we introduce a syntactic signature TODO where
different terms are only allowed to yield the same string whenever they
represent different syntactic structures of a syntactically ambiguous
phrase. This means that purely semantic distinctions such as scope
ambiguity have to be moved to a more abstract signature from whose terms
the semantic representation can be derived by a function. This approach
is studied more deeply in \cite{pogodalla2007generalizing}.

Another interesting composition pattern is to take an existing structure
of grammars and add a new abstract signature on top. This signature can
be used to further constrain the items in (TODO: what was before the
abstract-most signature/language). In \cite{pogodalla2012controlling},
this is what the authors use to develop their constraints on
extraction. Each of their constraints is presented as a new abstract
signature constraining the original syntactic signature.

However, a question that the authors do not pose is how to combine all
these constraints. If we were to apply the same pattern repeatedly to
get a structure like the one on (TODO: INSERT A DIAGRAM HERE), we would
have to deal with successively more complex type systems on each
level. This is due to the condition that a lexicon has to be a
homomorphism and thus it must, among other things, always map well-typed
terms to well-typed terms. To illustrate the problem, imagine a
structure such as the one on (TODO) that handles some different
constraints in each of the vertically stacked signatures including
e.g. agreement. If we tried to add a new abstract signature at the top
to implement another constraint, we would have to ensure that all the
terms we licence in our new abstract signature have a well-typed
counterpart in the existing abstract language. However, this means that
our new signature must make sure that the terms it approves do not break
agreement or any of the previously established constraints. Effectively,
our new signature would be forced to reimplement all of the existing
constraints, rendering vain all our effort to split up the constraints
into multiple signatures.

What we would like instead is to put the signatures side-by-side such
that they do not depend on each other (as on diagram TODO). However, how
does this actually define a language? We can arrive at the answer by
examining how are the graphical representations of ACGs connected to the
languages they define. We will associate a language to every node of the
graph. Before we begin, we notice that every ACG diagram formed by the
above rules of composition is a directed tree and therefore every node
has at most one parent/predecessor. To the root of the tree, the
abstract-most signature in the graph, we will assign the abstract
language generated by that signature and its distinguished type. To all
the other nodes, we will assign the result of mapping their parent's
language by the lexicon linking the two. By homomorphism of the
lexicons, we can also deduce the distinguished types for all the
non-root nodes. They are the images of their parents' distinguished
types by the connecting lexicons. For all of the nodes, it is true that
every term in the language assigned to the node has the node's
distinguished type.

We can confirm that the above graphical model of defining ACGs is
coherent. First, we can show that our interpretation is sound and so
that any directed tree adorned with signatures on the nodes, compatible
lexicons on the edges and a distinguished type on the root can only
define ACG languages. This is trivial, since the language corresponding
to the root is an ACG abstract language and the class of ACG languages
is closed under transformation by a lexicon and therefore all of the
languages corresponding to the descendants of the root are ACG languages
as well. Second, we can try and show some kind of completeness. However,
there is not much formalization done to describe a ``system of ACGs that
share common signatures'' (something that our generalization will
fix). We can nevertheless show that it is possible to express any ACG
using our graphical model. Given a grammar $\mathcal{G} =
\mathopen{<}\Sigma_A, \Sigma_O, \mathcal{L}, S\mathclose{>}$, we just
produce a tree whose root has the signature $\Sigma_A$ and distinguished
type $S$ and which is connected to a single child node having the
signature $\Sigma_O$ by the lexicon $\mathcal{L}$. Then the two nodes of
our tree define both the abstract and the object language defined by
$\mathcal{G}$.

We have now given some formal account about how we can read language
definitions in diagrams. Still, we have kept our analysis to the tree
diagrams used in current ACG literature, where every node has at most
one parent. To make the generalization more visible, we will switch from
an algebraic manner of presentation, in which we stated that the
language of a node is equal to an image of its parent's language, to a
calculus-based manner of presentation, in which we will state that the
language of a node is the set of all terms that have an antecedent in
the parent's language. Besides this property, there is also the fact
that every term in a node's language has the node's distinguished
type. This property is trivially true for nodes with parents because of
the homomorphic property of lexicons. However, it is crucial for the
root node for which this is the single defining property of its
language. To sum up, we have that for every node of a tree diagram, its
language is the set of all terms that have the node's distinguished type
and that have an antecedent in the parent's language, if there is
any. This rule is already general enough to handle the case of zero or
one parents, we can easily generalize it to handle an arbitrary count:
\emph{the language of a node is the set of all terms that have the
  node's distinguished type and that have antecedents in the languages
  of all of the node's parents}. If we transpose this back to the
algebraic style of presentation, we say that the language of a node is
the intersection of the set of terms having the node's distinguished
type and the intersection of all its parents' languages.

TODO: In the previous paragraph, add a notice saying that the terms must
be also built on the signature of the node.

We will now turn to a formal definition of this generalized graph-based
formalism and discuss some its values and properties.
